Update and evaluation of the ozone dry deposition in the Oslo~CTM3 v1.0

### 2019-07-04 DryDep.tex
Finished sec 3.2.1 and updated one plot.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparison with modeling results}
\label{subsec:model}
%
In our evaluation of results, we closely follow suggestions by \citet{ACP:Hardacre2015}. For the purpose of comparison with the multi-model mean of the therein participating Task Force on Hemispheric Transport of Air Pollution (TF~HTAP) models, we also have re-gridded our data to a horizontal resolution of $3^\circ\times 3^\circ$. In Section~\ref{subsubsec:zonal}, we look at zonal distributions of $[\chem{O_3}](z_0)$, $v^\chem{O_3}_\text{DD}$, and $\sum\chem{O_3^\text{DD}}$ for all our sensitivity simulations and study seasonal cycles of hemispheric ozone as well as for nine land surface types which have been retrospectively separated (Section~\ref{subsubsec:seasons}). From this, we estimate the total annual ozone dry deposition onto ocean, ice, and land surfaces and compare also with results from \citet{ACP:Luhar2017}.

As dry deposition velocities are not directly available from the model output, monthly average dry deposition velocities $v^\chem{O_3}_\text{DD}$ have been reconstructed from the ratio between the total ozone dry deposition $\sum \chem{O_3^\text{DD}}(z_0)$ and monthly averaged surface ozone amount $\chem{O_3}(z_0)$ 
\begin{equation}
  v^\chem{O_3}_\text{DD} = \frac{\sum \chem{O_3^\text{DD}}(z_0)}{\chem{O_3}(z_0)}\cdot c_\text{month}.
\end{equation}
Herein, $c_\text{month} = \frac{\Delta h_\text{month}}{s_\text{month}}$, with the monthly average height of the lowermost model level in each gridbox $\Delta h_\text{month}$ and the respective number of seconds in a month $s_\text{month}$.
%
\subsubsection{Zonal distribution}
\label{subsubsec:zonal}
%
The annual zonal average with respect to surface ozone concentration (Fig.~\ref{fig:mmm_drydep}a displays on average, in consistency with Fig.~\ref{fig:diff_maps}a, a global increase of surface ozone concentrations by up to $10\,\unit{ppb}$ when applying the mOSaic scheme compared to the Wesely scheme. This increase is largest in the zonal band $(25-50)\,\unit{^\circ N}$ which contains the major deserts. In the deep tropics ($5\,\unit{^\circ S}-5\,\unit{^\circ N}$), the increase is smallest ($\Delta[\chem{O_3}]\leq 5\,\unit{ppb}$). We find that the mOSaic scheme further intensifies the strong asymmetry between northern and southern hemisphere as a consequence of the distribution of the continental land masses and vegetation thereon. Among the sensitivity studies, there is only a low absolute variance. Most remarkable, but expected due to the much smaller prescribed dry deposition velocity over ice and snow, \emph{EMEP\_ppgssh\_ice} displays almost a doubling of surface ozone in the high Arctics compared to \emph{Wesely\_type} but affects ozone concentrations down to latitudes at about $50\,\unit{^\circ}$ in both hemispheres. Neglecting any dependence on solar radiation in the stomatal conductance formulation (\emph{EMEP\_offLight}) -- or in other words allowing photosynthesis 24/7 -- generally decreases the ozone concentration by about $2\,\unit{ppb}$, while choosing soil water at deeper levels than the surface layer (\emph{EMEP\_SWVL4}) increases $[\chem{O_3}](z_0)$ slightly. Rather surprisingly, switching off the phenology completely (\emph{EMEP\_offPhen}), amounts only to small difference (at most $3.5\,\unit{\%}$ increase in the deep tropics). Switching from the SMOKE-BEIS parameterization to the pre-processed greening season (\emph{EMEP\_ppgs} compared to \emph{EMEP\_full}), surface ozone increases on average by $6\,\unit{\%}$ between $(54-70)\,\unit{^\circ N}$. The scenario of differing emissions (2005 in comparison to 2014 or more specifically \emph{EMEP\_ppgs\_2005} compared to \emph{EMEP\_ppgs}), yields higher ozone concentrations in the northern hemisphere in 2005 in accordance to a reduction in sulfur and \chem{NO_x} emissions in south east Asia in later years. An opposite tendency is seen for latitudes south of $30\,\unit{^\circ N}$, probably owing to the industrial development of countries in these regions.

The $v^\chem{O_3}_\text{DD}$ are shown in Fig.~\ref{fig:mmm_drydep}b. The dry deposition velocities in the mOSaic scheme are on average $43\,\unit{\%}$ lower than in the Wesely scheme. In the Arctics, except for \emph{EMEP\_ppgssh\_ice}, all realizations of the mOSaic scheme are constantly above the multi-model-mean. This indicates, that with respect to other models, the updated dry deposition velocity above ice and snow should be considered as new standard for the Oslo~CTM3. This may, however, lead to an overcompensation of the current low-bias in surface ozone in the Oslo~CTM3. The dry deposition velocities are of course independent of the emission scenario, but display a strong dependence on $f_\text{light}$ and $f_\text{phen}$. The closer to $1$ these factors are the higher the dry deposition velocities become (compare \emph{EMEP\_offLight}, \emph{EMEP\_offPhen}, and \emph{EMEP\_ppgssh}). That highlights the importance of a proper PAR from meteorological input in CTMs or radiative transfer in PAR in global climate models (GCMs) and a proper choice of phenology for canopy conductance computation. Albeit, the total numbers are smaller, the shape of the normalized zonal average dry deposition velocities of the mOSaic scheme and the multi-model-mean are similar (Supplement~S.6: Fig.~S8). The biggest exceptions are the zonal bands $(50-70)\,\unit{^\circ S}$, which is almost entirely covered by ocean, $(12-30)\,\unit{^\circ S}$, which coincides with the location of Australia and its desert regions, as well as its counterpart in the northern hemisphere $(12-30)\,\unit{^\circ N}$. However, it is not clear whether diverging dry deposition velocities over ocean or deserts have the larger impact.

The annual total ozone dry deposition is shown in Fig.~\ref{fig:mmm_drydep}c. In accordance to the previously described features, we observe a reduction of the global total ozone dry deposition in all sensitivity studies to almost one-half of the amount given by \emph{Wesely\_type}. The occurrence of this reduction in the zonal bands, where the major deserts are located, points to a change in $v_\text{desert}^\chem{O_3}$. Consulting the parameter file used in the Wesely scheme, we indeed find $v_\text{desert}^\chem{O_3} \equiv v_\text{tundra}^\chem{O_3} = 0.26\,\unit{cm\,s^{-1}}$ \citep{JGR:Hough1991}, while in the mOSaic scheme $v_\text{desert}^\chem{O_3}=0.05\,\unit{cm\,s^{-1}}$ and $v^\chem{O_3}_\text{tundra} = 0.24\,\unit{cm\,s^{-1}}$, respectively. Similarly, dry deposition velocities over ice and snow and ocean have been even higher in the Wesely scheme ($v_\text{ice/snow}^\chem{O_3}\equiv v_\text{water}^\chem{O_3}=0.07\,\unit{cm\,s^{-1}}$) than in the original mOSaic scheme ($v_\text{ice/snow}^\chem{O_3}\equiv v_\text{water}^\chem{O_3}=0.05\,\unit{cm\,s^{-1}}$). These differences in baseline dry deposition velocities over huge parts of the unvegetated surface of the Earth amount to most of the quantitative difference between the Wesely and the mOSaic scheme. We further elaborate on this in the following section (Section~\ref{subsubsec:seasons}).

There seems to be a discrepancy between the Oslo~CTM3 response and the multi-model-mean, since the Wesely scheme is similar to the multi-model-mean with respect to total annual ozone dry deposition, while the reconstructed $v_\text{DD}^\chem{O_3}$ of the mOSaic scheme match better. This could be a sign of differences in photo-chemistry and transport (e.g. convective, advective, STE) between the Oslo~CTM3 and the average TF~HTAP model, but without comparing to the actual $[\chem{O_3}(z_0)]$ of the TF~HTAP models that participated in the model intercomparison, we cannot elaborate on this any further.

\begin{figure}[t]
  \includegraphics[width=8.3cm]{fig05}
  \caption{Comparison of the manifold of Oslo~CTM3 integrations with respect to (a) Surface ozone concentrations, (b) Annual average ozone dry deposition velocity, (c) Total annual ozone dry deposition. The different colors indicate sets of simulation with similar baselines. The multi-model mean from the evaluation of TF~HTAP models by \citet{ACP:Hardacre2015} is shown as a reference (where available).}
  \label{fig:mmm_drydep}
\end{figure}
%
In Appendix Fig.~\ref{fig:mmm_drydep_season}a, the average zonal ozone dry deposition is shown separated by month. Where available, we have added the multi-model-mean given by \citet{ACP:Hardacre2015} as reference. There are two major phases apparent: NH and SH greening season. Spring and summer in the NH is reflected in a pronounced peak of $v_\text{DD}^\chem{O_3}$ in the northern mid latitudes, while it is absent in winter (SH summer). Spring and summer in the SH are marked by a southward shift of the tropical peak dry deposition velocity and a slight increase of $v_\text{DD}^\chem{O_3}$ in the region $(20-40)\,\unit{^\circ S}$. In the Wesely scheme, NH mid latitude peak velocities appear in June compared to July in the mOSaic scheme, indicating that the seasonal cycles differ. The corresponding total monthly ozone dry deposition is shown in Appendix Fig.~\ref{fig:mmm_drydep_season}b. In general, the seasonal patterns are quite similar in the Wesely scheme and the mOSaic scheme, displaying a strong symmetry around $10\,\unit{^\circ N}$ in January/February and November/December, respectively. What differs most is the molding and intensity of the NH peak dry deposition. Both schemes reach the maximum in June/July but the peak is much more differentiated in March already in the Wesely scheme. Similarly, the SH tropical peak dry deposition is reached in August/September but sustained longer, into October, in the Wesely scheme. Since we have not conducted any simulation with a meteorological year other than 2005, we cannot elaborate on whether this is a special feature of our chosen year or not. In comparison, the available $2$ months of the multi-model-mean display a somewhat different pattern. Especially the NH/SH asymmetry is much more pronounced in February, than it is the mOSaic scheme.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Average seasonal cycles}
\label{subsubsec:seasons}
%
To further disentangle the contributions of different regions to the global ozone budget, we will briefly look at different projections of seasonal cycles.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparison with modeling results}
\label{subsec:model}
%
In the evaluation of our model, we closely follow suggestions by \citet{ACP:Hardacre2015}. For the purpose of comparison with the multi-model mean of the therein participating Task Force on Hemispheric Transport of Air Pollution (TF~HTAP) models, we also have re-gridded our data to a horizontal resolution of $3^\circ\times 3^\circ$. In Section~\ref{subsubsec:zonal}, we look at zonal distributions of $[\chem{O_3}](p_0)$, $v^\chem{O_3}_\text{DD}$, and $\sum\chem{O_3^\text{DD}}$ for all our sensitivity simulations and study seasonal cycles of hemispheric ozone as well as for nine land surface types (Section~\ref{subsubsec:seasons}). From this, we estimate the total annual ozone dry deposition onto ocean, ice, and land surfaces and compare also with results from \citet{ACP:Luhar2017}.

Dry deposition velocities are directly available only for the new model version. For \emph{Wesely\_type}, monthly average dry deposition velocities $v^\chem{O_3}_\text{DD}$ had to be retrospectively estimated from the ratio between the total ozone dry deposition $\sum \chem{O_3^\text{DD}}(z_0)$ and monthly averaged surface ozone amount $\chem{O_3}(z_0)$ 
\begin{equation}
  v^\chem{O_3}_\text{DD} = \frac{\sum \chem{O_3^\text{DD}}(z_0)}{\chem{O_3}(z_0)}\cdot c_\text{month}.
\end{equation}
Herein, $c_\text{month} = \frac{\Delta h_\text{month}}{s_\text{month}}$, with the monthly average height of the lowermost model level in each gridbox $\Delta h_\text{month}$ and the respective number of seconds in a month $s_\text{month}$.
%
\subsubsection{Zonal distribution}
\label{subsubsec:zonal}
%
The annual zonal average with respect to surface ozone concentration (Fig.~\ref{fig:mmm_drydep}a) displays on average, in consistency with Fig.~\ref{fig:diff_maps}a, a global increase of surface ozone concentrations by $14-15\,\unit{ppb}$ comparing \emph{mOSaic} to \emph{Wesely\_type}. This increase is largest in the zonal band $(25-50)\,\unit{^\circ N}$ which contains the major deserts. In the deep tropics ($5\,\unit{^\circ S}-5\,\unit{^\circ N}$), the increase is smallest ($\mathcal{O}(5\,\unit{ppb})$). We find that the mOSaic scheme further intensifies the strong asymmetry between northern and southern hemisphere as a consequence of the distribution of the continental land masses and vegetation thereon. Among the sensitivity studies, there is only a low absolute variance. Neglecting the dependence on light in the stomatal conductance formulation (\emph{mOSaic\_offLight}) -- or in other words allowing photosynthesis 24/7 -- decreases the ozone concentration by $1-2\,\unit{ppb}$ in the tropics and NH mid latitudes, while choosing soil water at shallower depths (\emph{mOSaic\_SWVL1}) increases $[\chem{O_3}]$ insignificantly. Rather surprisingly, switching off the phenology completely (\emph{mOSaic\_offPhen}), amounts on average only to small difference ($\mathcal{O}(<1\,\unit{ppb})$). Most remarkable, but expected due to the much smaller prescribed dry deposition velocity over ice and snow, \emph{mOSaic\_ice} displays a doubling of surface ozone in the high Arctics compared to \emph{Wesely\_type} ($\mathcal{O}(20\,\unit{ppb})$) but affects ozone concentrations down to latitudes at about $50\,\unit{^\circ}$ in both hemispheres. Reducing $R^\chem{O_3}_\text{desert}$ by $60\,\unit{\%}$ (\emph{mOSaic\_desert}) a reduction in the order of $1\,\unit{ppb}$ is found mainly limited to the NH. The largest impact on ozone concentrations ($\mathcal{O}(2-5\,\unit{ppb})$) is found for the experiment \emph{mOSaic\_hough} which is closest to \emph{Wesely\_type}, since we used on average the same $R^\chem{O_3}$ (see Supplement~S.7). The scenario of differing emissions (2005 in comparison to 2014 or more specifically \emph{mOSaic} compared to \emph{mOSaic\_emis2014}), yields higher ozone concentrations in the northern hemisphere in 2005 in accordance to a reduction in sulfur and \chem{NO_x} emissions in south east Asia in later years. An opposite tendency is seen for latitudes south of $30\,\unit{^\circ N}$, where an increase in ozone precursors is seen in CEDS.

The $v^\chem{O_3}_\text{DD}$ are shown in Fig.~\ref{fig:mmm_drydep}b. The dry deposition velocities in the mOSaic scheme are well below the Wesely scheme and in remarkable agreement with the results shown by \citet{ACP:Hardacre2015}. In the Arctics, except for \emph{mOSaic\_ice}, all model experiments are slightly above the multi-model-mean. This indicates, that with respect to the other models, the \citet{ACP:Helmig2007} dry deposition velocity above ice and snow should be considered as new standard for the Oslo~CTM3. This may, however, lead to an overcompensation of the current low-bias in surface ozone in the Oslo~CTM3 and needs further evaluation. The dry deposition velocities are of course independent of the emission scenario, but display a strong sensitivity on $f_\text{light}$ ($\mathcal{O}$), $f_\text{phen}$, and especially on the choice of $R^\chem{O_3}$. The shape of the normalized zonal average dry deposition velocities of the mOSaic scheme and the multi-model-mean are more similar than the \emph{Wesely\_type} (Supplement~S.6: Fig.~S8). The biggest exceptions are the zonal bands $(50-70)\,\unit{^\circ S}$, which is almost entirely covered by ocean, $(12-30)\,\unit{^\circ S}$, which coincides with the location of Australia and its desert regions, as well as its counterpart in the northern hemisphere $(12-30)\,\unit{^\circ N}$. However, it is not clear whether diverging dry deposition velocities over ocean or deserts are responsible for this.

The annual total ozone dry deposition is shown in Fig.~\ref{fig:mmm_drydep}c. In accordance to the previously described features, we observe a reduction of the global total ozone dry deposition in all sensitivity studies. In the most extreme case (NH subtropics and mid latitudes), the total ozone dry deposition drops to one-half of the amount given by \emph{Wesely\_type}. The occurrence of this reduction in the zonal bands, where the major deserts are located, points to a substantial difference in $v_\text{desert}^\chem{O_3}$. Consulting the parameter file used in the Wesely scheme, we indeed find $v_\text{desert}^\chem{O_3} \equiv v_\text{tundra}^\chem{O_3} = 0.26\,\unit{cm\,s^{-1}}$ \citep{JGR:Hough1991}, while in the mOSaic scheme $v_\text{desert}^\chem{O_3}=0.05\,\unit{cm\,s^{-1}}$ and $v^\chem{O_3}_\text{tundra} = 0.24\,\unit{cm\,s^{-1}}$, respectively. Similarly, dry deposition velocities over ice and snow and ocean have been even higher in the Wesely scheme ($v_\text{ice/snow}^\chem{O_3}\equiv v_\text{water}^\chem{O_3}=0.07\,\unit{cm\,s^{-1}}$) than in the original mOSaic scheme ($v_\text{ice/snow}^\chem{O_3}\equiv v_\text{water}^\chem{O_3}=0.05\,\unit{cm\,s^{-1}}$). These differences in surface dry deposition velocities over huge parts of the unvegetated surface of the Earth accounts for most of the qualitative difference between the Wesely and the mOSaic scheme, but does not explain the quantitative difference. We further elaborate on this in the following section (Section~\ref{subsubsec:seasons}).

There seems to be a discrepancy between the Oslo~CTM3 response and the multi-model-mean, since the Wesely scheme is similar to the multi-model-mean with respect to total annual ozone dry deposition, while the $v_\text{DD}^\chem{O_3}$ of the mOSaic scheme match better. This could be a sign of differences in photo-chemistry and transport (e.g. convective, advective, STE) between the Oslo~CTM3 and the average TF~HTAP model, but without comparing to the actual $[\chem{O_3}(p_0)]$ of the TF~HTAP models that participated in the model intercomparison, we cannot elaborate on this any further. Another hint may be found in the routines in the Oslo~CTM3 that actually compute the removal of ozone from the atmosphere which might have been adjusted to the less physical dry deposition velocities in the past but this subject to further investigations.

\begin{figure}[t]
  \includegraphics[width=8.3cm]{fig05}
  \caption{Comparison of the manifold of Oslo~CTM3 integrations with respect to (a) Surface ozone concentrations, (b) Annual average ozone dry deposition velocity, (c) Total annual ozone dry deposition. The different colors indicate sets of simulation with similar baselines. The multi-model mean from the evaluation of TF~HTAP models by \citet{ACP:Hardacre2015} is shown as a reference (where available).}
  \label{fig:mmm_drydep}
\end{figure}
%
In Appendix Fig.~\ref{fig:mmm_drydep_season}a, the average zonal ozone dry deposition is shown separated by month. Where available, we have added the multi-model-mean given by \citet{ACP:Hardacre2015} as reference. As for the global annual comparisons above, the mOSaic scheme matches the multi-model-mean values remarkably well with respect to dry deposition velocities, while it strongly underestimates the total dry deposition. There are two major phases apparent: NH and SH greening season. Spring and summer in the NH is reflected in a pronounced peak of $v_\text{DD}^\chem{O_3}$ in the northern mid latitudes, while it is absent in winter (SH summer). Spring and summer in the SH are marked by a southward shift of the tropical peak dry deposition velocity and a slight increase of $v_\text{DD}^\chem{O_3}$ in the region $(20-40)\,\unit{^\circ S}$. In the Wesely scheme, NH mid latitude peak velocities appear in June compared to July in the mOSaic scheme, indicating that the seasonal cycles differ. The corresponding total monthly ozone dry deposition is shown in Appendix Fig.~\ref{fig:mmm_drydep_season}b. In general, the seasonal patterns are quite similar in the Wesely scheme and the mOSaic scheme, displaying a strong symmetry around $10\,\unit{^\circ N}$ in January/February and November/December, respectively. What differs most is the molding and intensity of the NH peak dry deposition. Both schemes reach the maximum in June/July but the peak is much more differentiated in March already in the Wesely scheme. Similarly, the SH tropical peak dry deposition is reached in August/September but sustained longer, into October, in the Wesely scheme. Since we have not conducted any simulation with a meteorological year other than 2005, we cannot elaborate on whether this is a special feature of our chosen year or not.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Average seasonal cycles}
\label{subsubsec:seasons}
%
To further disentangle the contributions of different regions to the global ozone budget, we will briefly look at different projections of seasonal cycles.
